---
title: "YawPitchRollCalculation"
format:
    html:
      fig-width: 8
      fig-height: 4
      code-fold: true
author: Oliver Dürr
editor: visual
---

```{r setup, include=FALSE}
  library(tidyverse)
  library(onion) #For quaterions
```

This document describes how the orientation of the IMU in terms of yaw,pitch, and roll is calculated from the information provided by the IMU (we use the [MPU-9250](https://invensense.tdk.com/products/motion-tracking/9-axis/mpu-9250/)).

# Calculations as done on the Arduino

The result of the calculations in the Digital Motion Processor (DMP) done in the IMU are reported as quaternions (basically a short coding of the rotation between the IMU and the fixed laboratory system). Technically they are send to the Aduino (as a byte vector the `fifoBuffer`). On the Aduino the following steps happen:

1.  The result of the FIFO Buffer is parsed and written in a Quaternion with components $q=(q_w, q_x, q_y, q_z)$ this is done in \`dmpGetQuaternion\`

2.  The vector of gravity is calculated in `dmpGetGravity`

3.  From that vector (and the gravity) the yaw, pitch, and roll angles are calculated \`dmpGetYawPitchRoll\`

```{c, eval=FALSE, class.source = "fold-show"}
 dmpMpu[num].dmpGetQuaternion(&q, fifoBuffer); 
 dmpMpu[num].dmpGetGravity(&gravity, &q);
 dmpMpu[num].dmpGetYawPitchRoll(ypr, &q, &gravity);
```

## Getting the quaternion q

In the first line the quaternion send from the DSP is read from the `fifoBuffer` and parsed into the variable `q`. The quaternion `q` contains all information about a single IMU. This is done with the following code:

```{c, eval=FALSE}
uint8_t MPU6050::dmpGetQuaternion(int16_t *data, const uint8_t* packet) {
    // TODO: accommodate different arrangements of sent data (ONLY default supported now)
    if (packet == 0) packet = dmpPacketBuffer;
    data[0] = ((packet[0] << 8) | packet[1]);
    data[1] = ((packet[4] << 8) | packet[5]);
    data[2] = ((packet[8] << 8) | packet[9]);
    data[3] = ((packet[12] << 8) | packet[13]);
    return 0;
}

uint8_t MPU6050::dmpGetQuaternion(Quaternion *q, const uint8_t* packet) {
    // TODO: accommodate different arrangements of sent data (ONLY default supported now)
    int16_t qI[4];
    uint8_t status = dmpGetQuaternion(qI, packet);
    if (status == 0) {
        q -> w = (float)qI[0] / 16384.0f;
        q -> x = (float)qI[1] / 16384.0f;
        q -> y = (float)qI[2] / 16384.0f;
        q -> z = (float)qI[3] / 16384.0f;
        return 0;
    }
    return status; // int16 return value, indicates error if this line is reached
}
```

The yaw, pitch, roll information is calculated from the quaternion in two steps. First the vector pointing to the direction of gravity is calculated in `dmpGetGravity`

## Calculation of the vector of gravity

This is the code from MPU6050_6Axis_MotionApps20.h

```{c, eval=FALSE}
        uint8_t MPU6050::dmpGetGravity(VectorFloat *v, Quaternion *q) {
          v -> x = 2 * (q -> x*q -> z - q -> w*q -> y); 
          v -> y = 2 * (q -> w*q -> x + q -> y*q -> z);
          v -> z = q -> w*q -> w - q -> x*q -> x - q -> y*q -> y + q -> z*q -> z;
          return 0;
        }
```

Mathematically speaking this code constructs the vector of gravity as follows:

$$
\begin{align}
  v_x &= 2 (q_x q_z - q_w q_y) \\
  v_y &= 2 (q_w q_x + q_y q_z) \\
  v_z &= q_w^2 - q_x^2 - q_y^2 + q_z^2
\end{align}
$$

This translates into the following R code:

```{r}
dmpGetGravity = function(q){
  qw = q[1]
  qx = q[2]
  qy = q[3]
  qz = q[4]
  vx = 2 * (qx * qz - qw*qy)
  vy = 2 * (qw * qx + qy*qz)
  vz = qw^2 - qx^2 - qy^2 + qz^2
  return(c(vx,vy,vz))
}
q.vec = as.matrix(c(0.32,  0.30,  0.29, -0.85), ncol=1)
dmpGetGravity(q.vec)
```

### What the code does

What is happening in the code? We are asking "How does the gravity" looks like on the IMU $g_{\tt{IMU}}$? The gravity in the laboratory frame is given by \$g\_{\\tt{lab}}=(0,0,1)\$. Note that, we could also define a different lab frame in which the gravity is e.q. given by $(0,1,0)$. From the Quaternion Algebra (see e.g. [Note_on_Quaternion](https://oduerr.github.io/gesture/Note_on_Quaternion.html)), we know that:

$$
 q_{\tt{lab}} = q \; q_{\tt{IMU}} \; q^{-1}
$$

Where $q_{\tt{lab}}=(0,0,0,1)$ and $q_{\tt{IMU}}=(0,g_x, g_y, g_z)$ are pure quaternions, i.e. they take the vectors and set the real part $w$ to 0. To solve for $q_{IMU}$, we multiply with $q^{-1}$ and $q$ from left and right. Getting:

$$
  q_{\tt{IMU}}  = q^{-1} \; q_{\tt{lab}} \; q
$$ We can verify this with the R-Code:

```{r}
q = as.quaternion(q.vec) 
qz = quaternion(Re = 0, k = 1)
qz
q^-1 * qz * q
gravity = dmpGetGravity(q.vec)
gravity
```

## Calculation of yaw, pitch, roll

The code in `MPU6050_6Axis_MotionApps20.h` used looks like follows:

```{c, eval=FALSE}
uint8_t MPU6050::dmpGetYawPitchRoll(float *data, Quaternion *q, VectorFloat *gravity) {
  // yaw: (about Z axis)
  data[0] = atan2(2*q -> x*q -> y - 2*q -> w*q -> z, 2*q -> w*q -> w + 2*q -> x*q -> x - 1);
  // pitch: (nose up/down, about Y axis)
  data[1] = atan(gravity -> x / sqrt(gravity -> y*gravity -> y + gravity -> z*gravity -> z));
  // roll: (tilt left/right, about X axis)
  data[2] = atan(gravity -> y / sqrt(gravity -> x*gravity -> x + gravity -> z*gravity -> z));
  return 0;
}
```

In math:

\$\$ \begin{align}
  yaw &= {\tt atan2} (2 q_x q_y - 2 q_w q_z, 2 q_w^2 + 2 q_x^2 -1) \\
  
  pitch &= {\tt atan} (\frac{g_x}{\sqrt{g_y^2 + g_z ^2}}) \\
  roll &= {\tt atan} (\frac{g_y}{\sqrt{g_x^2 + g_z ^2}} )
\end{align} \$\$

This translate to the following R-Code.

```{r}
dmpGetYawPitchRoll = function(q, gravity){
  qw = q[1]
  qx = q[2]
  qy = q[3]
  qz = q[4]
  yaw = atan2(2*qx*qy - 2*qw*qz, 2*qw*qw+2*qx*qx-1)
  pitch = atan(gravity[1]/sqrt(gravity[2]^2+gravity[3]^2))
  roll = atan(gravity[2]/sqrt(gravity[1]^2+gravity[3]^2))
  return(c(yaw, pitch, roll))
}
as_arduino = dmpGetYawPitchRoll(q@x, gravity)
round(as_arduino,2)
```

The result from the Arduino is: (2.28, -0.77, -0.31). We can reproduce the results from the Arduino (see Comparison).

### What the code does

For yaw it extracts the angle using quaternions. For pitch it uses the before calculated gravity (in the IMU space). Let's start with the pitch and roll. From https://ahrs.readthedocs.io/en/latest/filters/tilt.html we find (with $a$ being the normalized gravity vector in the IMU space).

$$
  {\tt pitch} = \arctan2(-a_x, \sqrt{a_y^2 + a_z^2}) \\
  {\tt roll} = \arctan2(a_y, a_z) 
$$ Pitch:

```{r}
#Pitch
atan(gravity[1] / sqrt(gravity[2]^2 + gravity[3]^2)) #As done in Aduino
#
a = gravity/sqrt(sum(gravity^2)) #We are normalized
atan2(-a[1], sqrt(a[2]^2 + a[3]^2)) 
```

Except the sign, we got the same value as the code on the Arduino.

Roll:

```{r}
#Roll
atan(gravity[2] / sqrt(gravity[1]^2 + gravity[3]^2)) #As done in Arduino
atan2(gravity[2], gravity[3])
```

For the roll, we get a different value compared to the Arduino. 

# Comparison with other methods

Usually the YPR is calculated differently.

```{r}
  #From poyu (value9250.py)
  #def quart_to_rpy(w, x, y, z):
  #  roll = math.atan2(2 * (w * x + y * z), 1 - 2 * (x * x + y * y))
  #  pitch = math.asin(2 * (w * y - x * z))
  #  yaw = math.atan2(2 * (w * z + x * y), 1 - 2 * (z * z + y * y))
  #  return yaw, pitch, roll
  
  quart_to_ypr_value9250 = function(q){
    w = q[1]
    x = q[2]
    y = q[3]
    z = q[4]
    roll = atan2(2 * (w * x + y * z), 1 - 2 * (x * x + y * y))
    pitch = asin(2 * (w * y - x * z))
    yaw = atan2(2 * (w * z + x * y), 1 - 2 * (z * z + y * y))
    return(c(yaw,pitch,roll))
  }
  
  #https://stackoverflow.com/questions/5782658/extracting-yaw-from-a-quaternion
# This function has yaw and roll mixed put (Euler, vs Aircarft) 
  if (FALSE){
    quart_to_ypr_stack = function(q){
    yaw = atan2(2.0*(q[3]*q[4] + q[1]*q[2]), q[1]*q[1] - q[2]*q[2] - q[3]*q[3] + q[4]*q[4])
    pitch = asin(-2.0*(q[2]*q[4] - q[1]*q[3]))
    roll = atan2(2.0*(q[2]*q[3] + q[1]*q[4]), q[1]*q[1] + q[2]*q[2] - q[3]*q[3] - q[4]*q[4])
      return(c(yaw,pitch,roll))
  }
  }
  
 quart_to_ypr_chatgpt = function(q) {
    # Import necessary libraries
    w = q[1]
    x = q[2]
    y = q[3]
    z = q[4]

    # Convert the quaternion to Euler angles
    roll = atan2(2 * (w * x + y * z), 1 - 2 * (x * x + y * y))
    pitch = asin(2 * (w * y - z * x))
    yaw = atan2(2 * (w * z + x * y), 1 - 2 * (y * y + z * z))
    return(c(yaw, pitch, roll))
 }

  as_arduino
  quart_to_ypr_value9250(as.numeric(q))
  #quart_to_ypr_stack(as.numeric(q)) #Here roll and yaw are swaped
  quart_to_ypr_chatgpt(as.numeric(q))
```

Testing the routines, for different values. See https://matthew-brett.github.io/transforms3d/reference/transforms3d.euler.html
```{r}
  q = quaternion(Re = 0.99810947, i=0.06146124)
  quart_to_ypr_value9250(as.numeric(q))
  quart_to_ypr_chatgpt(as.numeric(q))
```



# Comparison

This code shows that, we can reproduce the calculation done on the Adunio.

```{r, eval=TRUE,show=FALSE}
  training <- read.csv("../../data/value06.csv",header=TRUE)
  training = rename(training, IMU=X)
  
  training$IMU = as.factor(training$IMU)
  nrow(training)
  training$t=rep(1:(1+nrow(training)/6), each=6)[1:nrow(training)]
  
  dat = training %>% filter(IMU != "0")
  dat_q = dat %>% select(starts_with('q'))

  ypr_calc = matrix(NA, ncol=3, nrow = nrow(dat_q))
  for (i in 1:nrow(dat_q)){
    gr = dmpGetGravity(t(dat_q[i,]))
    ypr_calc[i,] = dmpGetYawPitchRoll(t(dat_q[i,]), gr)
  }
  r2d = 360/(2*pi)
  plot(r2d*ypr_calc[,1], r2d*dat$yaw, main='Calculated YAW (R vs Aduino) in degrees')
  plot(r2d*ypr_calc[,2], r2d*dat$pitch, main='Calculated Pitch (R vs Aduino) in degress')
  plot(r2d*ypr_calc[,3], r2d*dat$roll, main='Calculated Roll (R vs Aduino) in degrees')
```

These plots show that, we can reproduce the Aduino values in R.
